Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-24 src/article.tex
Updates from Overleaf

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—typically during concerts or by experienced performers with special permission from the curator. To enhance accessibility and engagement, the museum commissioned the replica of a historical keyboard built in the tradition of the Italian Renaissance. The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. The interface is currently connected to a software sampler loaded with the Spitfire sample library, though the ultimate goal is to use samples of instruments from the collection. Visitors to the muesum are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of interface's construction and offers reflections on its role within the Tagliavini Collection and potential application within the wider musical instrument museum context.

@@

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

A modular system of printed circuit boards (PCBs) was designed to manage the sensors and process their output. The 49 QRE1113 sensors were distributed across seven boards, containing 7 sensors each. The PCBs were secured to the underside of the keyboard mechanism, allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

@@

\subsection{Project Deployment}
A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an \emph{open-source} approach for all hardware, software, and data. The commitment to open-sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

@@

The Arduino Nano BLE was chosen as the core microcontroller for its compatibility with open-source tools and ability to support both USB and BLE MIDI. Ferroelectric RAM (FRAM) was employed for data storage since it offers an economical yet robust solution for preserving calibration settings across power cycles. Calibration workflows were optimised using the Arduino IDE’s serial plotter and open-source MIDI Monitor software, reducing reliance on proprietary tools and simplifying the process for users.

Reference repositories for this project can be found here. FILL OUT AS APPROPRIATE (and anonymise).

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
    \\
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software, as per Figure \ref{fig:serial_monitor}. As an expert harpsichord performer, finer calibration was carried out with the guidance of \anon{Catalina Vicens}.

>>>>

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—typically during concerts or by experienced performers with special permission from the curator. To enhance accessibility and engagement, the museum commissioned the replica (Figure \ref{fig:teaser}) of a historical keyboard built in the tradition of the Italian Renaissance. The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. The interface is currently connected to a software sampler loaded with the Spitfire sample library, though the ultimate goal is to use samples of instruments from the collection. Visitors to the muesum are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of interface's construction and offers reflections on its role within the Tagliavini Collection and potential application within the wider musical instrument museum context.

@@

A modular system of printed circuit boards (PCBs) was designed to manage the sensors and process their output. The system uses 49 QRE1113 sensors which are distributed across seven boards, containing 7 sensors each. The PCBs were secured to the underside of the keyboard mechanism, allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

@@

\subsection{Project Deployment}
A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an \emph{open source} approach for all outputs of the project. The commitment to open sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{src/images/49-key-bottom-sensors-no-keys.jpg} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

@@

The Arduino Nano BLE was chosen as the core microcontroller for its compatibility with open source tools and ability to support both USB and BLE MIDI. Ferroelectric RAM (FRAM) was employed for data storage since it offers an economical yet robust solution for preserving calibration settings across power cycles. Calibration workflows were optimised using the Arduino IDE’s serial plotter and open source MIDI Monitor software, reducing reliance on proprietary tools and simplifying the process for users.

Reference repositories for this project can be found here:
\begin{anonsuppress}
    \begin{itemize}
        \item 
        \emph{Firmware}: \url{https://github.com/Nemus-Project/haprsichord-interface-firmware}
        \item 
        \emph{PCB CAD}: \url{https://github.com/Nemus-Project/haprsichord-interface-cad}
        \item 
        \emph{Interface Models}: \url{https://github.com/Nemus-Project/haprsichord-interface-models}
    \end{itemize}
\end{anonsuppress}
% FILL OUT AS APPROPRIATE (and anonymise).

@@

\begin{figure}[b]
    \centering
    \includegraphics[width=\linewidth]{src/images/CrossSectionSensorPlacement.jpg}
    \\
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open source MIDI monitoring software, as per Figure \ref{fig:serial_monitor}. As an expert harpsichord performer, finer calibration was carried out with the guidance of \anon{Catalina Vicens}.
